Modelling last glacial cycle ice dynamics in the Alps

### 2018-07-19 articles/cycle/response-referee1.tex

    The sensitivity of our results to the 1\,\% threshold was not tested.
    However, in our model results, liquid water fractions above 1\,\% typically
    only occur within the basal temperate layer of the fastest-moving glaciers
    where ice movement is dominated by basal sliding. We suspect that a local
    increase of ice deformation there is negligible in comparaison to
    uncertainties caused by basal sliding and more importantly climate forcing.

    We prefer to avoid including the above discussion in the manuscript, but we
    have reworked the sentence on water content:

    \msquote{%
        [Ice softness] increases with liquid water fractions up to 0.01
        \citep[p.~65--66]{Duval.1977, Lliboutry.Duval.1985,
        Cuffey.Paterson.2010}, an arbitrary threshold above which new
        measurements are critically needed \citep{Kleiner.etal.2015}.}

    The uncertainty to unknown rheology of water-rich temperate ice was also
    mentioned in the conclusions:

    \msquote{%
        A constant rheology was used for temperate ice containing more than
        1\,\% of liquid water.}

@@

    We refer to \citet{Bueler.Pelt.2015} as their paper contain the most
    up-to-date description of PISM till effective pressure physics used in our
    simulations \citet[Eqs.~18, 23, and 24]{Bueler.Pelt.2015}. However,
    subglacial water is not routed down the hydropotential gradient. We have
    clarified this:

    \msquote{%
        Effective pressure is related to the ice overburden stress and the
        modelled amount of subglacial water, using a~formula derived from
        laboratory experiments with till extracted from the base of Ice Stream
        B in West Antarctica \citep[Table~1;][Eqs~23 and
        24]{Tulaczyk.etal.2000, Bueler.Pelt.2015}. Basal meltwater is
        accumulated locally without transportation. When the till becomes
        saturated, additional meltwater assumed to drain off instantaneously
        outside the glacier margins, i.e. it is removed from the system in an
        accountable way.}

    \referee{\textbf{Somewhat related:}
        Little information is given on the choices of basal sliding parameter
        values in Table 1. This could be discussed briefly. Presumably no
        inversion or optimization was performed for these values beforehand,
        and they do vary spatially. Are they appropriate for Alpine bedrock
        overall?}

    In general, we chosed to not repeat parameter values from Table~1 in the
    main text. The following text was added in the methods:

    \msquote{%
        [a constant basal friction angle] corresponding to the average of
        available measurements \citep[p.~268]{Cuffey.Paterson.2010}. [...]
        Other parameters (Table~1) follow simulations of the Greenland ice
        sheet \citep{Aschwanden.etal.2013}, or benchmarks when other data is
        missing \citep{Bueler.Pelt.2015}.}

    Inversion of specific basal sliding parameters for past Alpine glaciers
    is difficult because the altitude and age of maximum ice surface elevation
    is discussed (cf. introduction and discussion on trimlines), and also
    depends on the even more uncertain regional climate history.  Therefore,
    basal sliding was also mentioned as one of the major sources of
    uncertainty in the conclusions.

    \msquote{%
        The till deformation model used here does not hold for sliding over
        bedrock surfaces. On the other hand, the constant friction angle used
        is representative of wet till but weaker basal conditions may have
        applied over saturated lake sediments where they occured.}

### 2018-07-19 articles/cycle/alps-cycle.tex
Assess referee 1 comment on basal sliding.

    A~pseudo-plastic sliding law relates the bed-parallel shear stresses to the
    sliding velocity. The yield stress is modelled using the Mohr--Coulomb
    criterion using a constant basal friction angle
    \citep[Table~\ref{tab:params};][Eq.~18]{Tulaczyk.etal.2000}.
    Effective pressure is related to the ice overburden stress and the modelled
    amount of subglacial water, using a~formula derived from laboratory
    experiments with till extracted from the base of Ice Stream B in West
    Antarctica \citep[Table~\ref{tab:params};][Eqs~23 and
    24]{Tulaczyk.etal.2000, Bueler.Pelt.2015}. Basal meltwater is accumulated
    locally without transportation. When the till becomes saturated, additional
    meltwater assumed to drain off instantaneously outside the glacier margins,
    i.e. it is removed from the system in an accountable way.

@@

    However, these results are limited by uncertainties on basal sliding and
    on the unknown rheology of water-rich temperate ice. The potential
    variability in surface mass balance parameters, as well as the simplicity
    of the climate forcing, are even more important.
    In particular, additional palaeo-climate variability over
    the European Alps may have caused more glaciations than modelled here.
    Using more specifically targeted sensitivity runs, future
    modelling studies will certainly be able to quantify uncertainties
    associated with some of the above limitations. Nevertheless, we hope that
    these conclusions will also serve as a basis for future studies of glacial
    geology in the Alps, and call for a more systematic aggregation and
    homogenization of glacial geological data to form a basis for model
    validation across the entire Alpine range.

>>>>

    A~pseudo-plastic sliding law relates the bed-parallel shear stresses to the
    sliding velocity. The yield stress is modelled using the Mohr--Coulomb
    criterion \citep[Table~\ref{tab:params};][Eq.~18]{Tulaczyk.etal.2000}.
    using a constant basal friction angle corresponding to the average of
    available measurements \citep[p.~268]{Cuffey.Paterson.2010}.
    Effective pressure is related to the ice overburden stress and the modelled
    amount of subglacial water, using a~formula derived from laboratory
    experiments with till extracted from the base of Ice Stream B in West
    Antarctica \citep[Table~\ref{tab:params};][Eqs~23 and
    24]{Tulaczyk.etal.2000, Bueler.Pelt.2015}. Basal meltwater is accumulated
    locally without transportation. When the till becomes saturated, additional
    meltwater assumed to drain off instantaneously outside the glacier margins,
    i.e. it is removed from the system in an accountable way. Other parameters
    (Table~\ref{tab:params}) follow simulations of the Greenland ice sheet
    \citep{Aschwanden.etal.2013}, or benchmarks when other data is missing
    \citep{Bueler.Pelt.2015}.

@@

    However, these results are limited by uncertainties on ice physics. The
    the till deformation model used here does not hold for sliding over
    bedrock surfaces. On the other hand, the constant friction angle used is
    representative of wet till but weaker basal conditions may have applied
    over saturated lake sediments where they occured. A constant rheology was
    used for temperature ice containing more than 1\,\% of liquid water.

    More importantly, these results are limited by the simplicity of the
    surface mass balance parameters and climate forcing used.
    In particular, additional palaeo-climate variability over
    the European Alps may have caused more glaciations than modelled here.
    Using more specifically targeted sensitivity runs, future
    modelling studies will certainly be able to quantify uncertainties
    associated with some of the above limitations. Nevertheless, we hope that
    these conclusions will also serve as a basis for future studies of glacial
    geology in the Alps, and call for a more systematic aggregation and
    homogenization of glacial geological data to form a basis for model
    validation across the entire Alpine range.
